Transfer learning between Sentinel-1 acquisition modes enhances \\ the few-shot segmentation of natural oil slicks in the Arctic

### 2025-04-03 latex/manuscript.tex
manuscript line numbering
manuscript orcid links
Added orcid links to author list

\documentclass[lettersize,journal]{IEEEtran}
\usepackage{amsmath,amsfonts}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{array}
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\usepackage{textcomp}
\usepackage{stfloats}
\usepackage{url}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}

% \usepackage{xcolor}
% \usepackage{soul}

@@

\author{Julien Vadnais,~\IEEEmembership{Graduate Student Member,~IEEE, }Benjamin Aubrey Robson, Christian Haug Eide, Rune Mattingsdal, Malin Johansson
        % <-this % stops a space
\thanks{Manuscript received March 31, 2025. This research was carried out in the context of the InFluSe project (Research Council of Norway Project 353044).}

@@

\begin{abstract} 
    Natural seepage is a significant contributor to marine hydrocarbon inputs. Remote and intermittent seeps are difficult to monitor in the field, yet oil slicks can be observed from spaceborne 
    synthetic aperture radar (SAR) due to differences in their backscatter,
    creating potential for automatic mapping. In mapping tasks such as segmentation, deep learning models excel, whilst needing large amounts of labeled images.
    To deal with scarcity of labeled images, transfer learning is an approach 
    commonly used in computer vision, yet still underutilized in remote sensing. In the case of oil slicks, differences between Sentinel-1 acquisition modes—such as the 
    interferometric wide (IW) in the North Sea and extra-wide (EW) in the Arctic—complicate direct model transfer.
    Here, we present a use-case where transfer learning enhances the segmentation of natural oil slicks. We used labeled slicks in IW images in the North Sea to pretrain a series of DeepLabv3 and SAM models. 
    These models were then fine-tuned on EW-labeled slicks from two documented Arctic seeps on which we have only limited observations. Our results show clear evidence that transfer learning improves few-shot 
    segmentation, even in challenging images with slick look-alikes. Overall, few studies have addressed transfer learning between SAR acquisition modes, and this work contributes to 
    better monitoring poorly understood or yet undiscovered natural oil seeps. 
\end{abstract}

\begin{IEEEkeywords}
    Natural seepage, oil slicks, SAR, deep learning, transfer learning, segmentation, DeepLab, SAM.
\end{IEEEkeywords}

@@

\section*{Acknowledgments}
Julien Vadnais received financial support from the Doctoral Research Scholarship of the Fonds de recherche du Québec. 
The authors would like to thank the EU Copernicus programme for allowing open access to the Sentinel-1 archives.
This research was carried out as part of the InFluSe project (Research Council of Norway Project 353044).
\bibliographystyle{IEEEtran}
\bibliography{references.bib}

>>>>

\documentclass[lettersize,journal]{IEEEtran}
\usepackage{amsmath,amsfonts}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{array}
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\usepackage{textcomp}
\usepackage{stfloats}
\usepackage{url}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[switch]{lineno}
\usepackage{orcidlink}
\usepackage{xcolor}

% ORCID macro
\newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\textcolor[HTML]{A6CE39}{\aiOrcid}}}
% \usepackage{soul}

@@

\author{
    Julien Vadnais\orcidlink{0009-0003-2680-8841},~\IEEEmembership{Graduate Student Member,~IEEE}, 
    Benjamin Aubrey Robson\orcidlink{0000-0002-4987-7378}, 
    Christian Haug Eide\orcidlink{0000-0003-4949-9917}, \\
    Rune Mattingsdal\orcidlink{0000-0003-0168-2405}, 
    Malin Johansson\orcidlink{0000-0003-0129-2239},~\IEEEmembership{Member,~IEEE} 
        % <-this % stops a space
\thanks{Manuscript received March 31, 2025. This research was carried out in the context of the InFluSe project (Research Council of Norway Project 353044).}

@@

\linenumbers
\begin{abstract} 
    Natural seepage is a significant contributor to marine hydrocarbon inputs. Remote and intermittent seeps are difficult to monitor in the field, yet oil slicks can be observed from spaceborne 
    synthetic aperture radar (SAR) due to differences in their backscatter,
    creating potential for automatic mapping. In mapping tasks such as segmentation, deep learning models excel, whilst needing large amounts of labeled images.
    To deal with scarcity of labeled images, transfer learning is an approach 
    commonly used in computer vision, yet still underutilized in remote sensing. In the case of oil slicks, differences between Sentinel-1 acquisition modes—such as the 
    interferometric wide (IW) in the North Sea and extra-wide (EW) in the Arctic—complicate direct model transfer.
    Here, we present a use-case where transfer learning enhances the segmentation of natural oil slicks. We used labeled slicks in IW images in the North Sea to pretrain a series of DeepLabv3 and SAM models. 
    These models were then fine-tuned on EW-labeled slicks from two documented Arctic seeps on which we have only limited observations. Our results show clear evidence that transfer learning improves few-shot 
    segmentation, even in challenging images with slick look-alikes. Overall, few studies have addressed transfer learning between SAR acquisition modes, and this work contributes to 
    better monitoring poorly understood or yet undiscovered natural oil seeps. 
\end{abstract}

\nolinenumbers
\begin{IEEEkeywords}
    Natural seepage, oil slicks, SAR, deep learning, transfer learning, segmentation, DeepLab, SAM.
\end{IEEEkeywords}
\linenumbers

@@

\nolinenumbers
\section*{Acknowledgments}
Julien Vadnais received financial support from the Doctoral Research Scholarship of the Fonds de recherche du Québec. 
The authors would like to thank the EU Copernicus programme for allowing open access to the Sentinel-1 archives.
This research was carried out as part of the InFluSe project (Research Council of Norway Project 353044).
\bibliographystyle{IEEEtran}
\bibliography{references.bib}
